### 2016-10-26 doc/publications/plos16/plos16.tex
changes to plos16.tex

Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs
(SNPs) by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  
Complex trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

\marginpar{decison trees capture interaction -- there is no main effect term. Random forests inherit this property. I dont think it
  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
prioritise SNPs for their contribution to disease.  The joint interaction is achieved by applying the technique of
bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
by summing the number of times a variable is used to split a node, over all nodes and all trees.  Random forests are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
developed in other programming languages.  {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast running
times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for
dealing with large data sets.

@@

\marginpar{Needs to be rewritten once we actually have the sections} 
In the first section we explore some of the
properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of random
forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
respect to the dimensionality of the data, building a random forest model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
\marginpar{what is ARI?}
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

\section{Methods}

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to useless predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

Random forests~\cite{Breiman2001} (RFs) apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data is independently drawn from the joint distribution of $(X,Y)$ and comprises $n$ $(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$.
$X$ and $\theta_b$ for $b=1,\ldots,B$ are i.i.d. random vectors and a tree is grown on each of these $\theta_b$ samples 
($X$ is randomly sampled $B$ times with replacement and a tree is grown on each of these $B$ samples). 
The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
\end{equation*}

@@

The average prediction error for an individual tree $h(X; \theta)$ is
\begin{equation}
PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
\end{equation}
Assume that, for all $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
\theta^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

@@

Since the original implementation of random forests, there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in the case where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.

@@

\subsubsection{convergence}
Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker.
\marginpar{say something about the rate of convergence}

Subject to certain conditions, random forests will not overfit even in the $p > n$ case. 
Here we demonstrate that this may hold true in extreme cases where $p \gg n$.

\marginpar{why is this relevant, i.e. if this is special to VariantSpark it should be explained better.}
\subsubsection{variable selection}
\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest.}
% DENIS: Aidan and Piotr can you please explain here how CursedForest is implemented
Traditionally, VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
in a ``DataFrame'', similar to DataFrames in R.

@@

The current implementation of \cursedforest\ uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labelled with value $q$ where $q = 1,
  \ldots, Q$ at a node. The entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable
\end{itemize}

@@

%DENIS: Rob how was the synthetic data generated
\subsubsection{Synthetic data} 
Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a two level
Bernoulli trail with $p$ (probability of success) calculated by applying inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.

>>>>

Identifying the relevant and disease specific base-pair difference between individuals is the focus of genome wide
association studies (GWAs).  These analysis are typically performed on only the most frequently differing base-pairs, termed
single nucleotide polymorphisms (SNPs), by applying linear or logistic regression analysis to each SNP separately~\cite{CCC2007}.  It has since been
demonstrated that not taking joint interactions between SNPs into account when filtering for potential disease mutations
is inadequate~\cite{Manolio2009}.  
Complex trait analysis has demonstrated that in order to explain the overall
heritability a model needs to take the joint contribution of all variants into account~\cite{Yang2011}.

%\marginpar{decison trees capture interaction -- there is no main effect term. Random forests inherit this property. I dont think it
%  has anything to do with boostrapping, the point of which is to ensure that successive trees are not correlated.}
%Random forests~\cite{Breiman2001} are able to capture interactions between features while still allowing to
%prioritise SNPs for their contribution to disease.  The joint interaction is achieved by applying the technique of
%bootstrap aggregation or bagging to decision trees, while a measure of variable importance can be provided, for example,
%by summing the number of times a variable is used to split a node, over all nodes and all trees.  
Random forests are well suited for
processing ``wide'' genomic data for two reasons.  Firstly, while other machine learning applications have the
propensity to overfit datasets with more features $p$ than samples $n$ (a consequence of the ``curse of
dimensionality''~\cite{Bauer2014, bellman1961adaptive}), decision trees are resistant to overfitting. 
Secondly, random forests are also very easy to parallelise. As the forest is a sum of decision trees, it it possible to grow
separate trees on different processors and combine the trees. See the discussion of the properties of random forests
Section \ref{section:methods}.

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}$), and the {\sc randomForest} implementation in R lending itself to parallelization to grow multiple trees
simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide'' genomic data have been
developed in other programming languages.  {\sc CloudForest}~\cite{Bressler2015} written in Go, achieves fast run
times by effective use of the CPU cache, optimising for different classes of features and efficiently multi-threading.
Similarly, {\sc Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is specifically optimised for
dealing with large data sets.

@@

In the first section we explore some of the
\marginpar{Needs to be rewritten once we actually have the sections} 
properties of random forests, demonstrating that any potential flaws in \cursedforest\ are an inherent property of random
forests. Here we provide demonstrations using synthetic data.  We then demonstrate the scalability of \cursedforest\ in
respect to the dimensionality of the data, building a random forest model on whole-genome data from the 1000 Genomes Project. Here
we report on the accuracy (in terms of the ARI)
\marginpar{what is ARI?}
 and the error.  Finally, given the role different parameter values can
play in model construction, we explore the effect tuning these parameters can have on the prediction accuracy of the
model.

\section{Methods}
\label{section:methods}

@@

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninforamtive predictors;
  \item are able to capture interactions between features (which is of importance for modelling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

Random forests  apply the technique of bootstrap aggregation or bagging to decision trees.  The training
data sets are independently drawn. with replacement,  from the joint distribution of $(X,Y)$ so that each set comprises $n$ $(p+1)$-tuples
 $(x_1,y_1),\ldots, (x_n,y_n)$. This si done $B$ times and  a tree is grown on each of these $B$ samples.
The results are combined in an appropriate way. In a classification problem we take the majority vote for the trees over
the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;\theta_b)=q)\right).
\end{equation*}

@@

The average prediction error for an individual tree $h(X; \theta)$ is
\marginpar{is this formulation correct in the case of classification? See Friedman on 0-1 losss}
\begin{equation}
PE_t = E_\theta E_{X,Y} (Y-h(X; \theta))^2.
\end{equation}
Assume that, for all $\theta$, the tree is unbiased, i.e., $EY= E_X h(X; \theta)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;\theta))$ and $(Y-h(X;\theta^\prime))$ for independent $\theta,
\theta^\prime$.  
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}. In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly selected subset of the variables.

@@

Since the original implementation of random forests, there have been a number of developments leading to a better understanding of
the algorithm. 
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in cases where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.

@@

Biau {\it et al.}~\cite{Biau.2012} provide a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the data set. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear waht the size
of teh set $S$ will be in a GWAs application. There are cases of small groups of SNPs with large effectsd and also of
contitions casued by the interaction of large groups of SNPs. 
\marginpar{check this}

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.
\marginpar{ \cite{Chen.and.Ishwaran.2012} cite many papers on extensions of random forests to handle 
GWAS data. What do we need to do in this area? What about linkage diseqelibrium?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest.}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. These algorithms require data
in a ``DataFrame'', similar to DataFrames in R.

@@

The current implementation of \cursedforest\ uses:
\begin{itemize}
\item an ``information gained'' criteria for splitting. Let $f_q$ be the fraction of items labelled with value $q$ where $q = 1,
  \ldots, Q$ at a node. The entropy of the node is $H = - \sum^{m}_{i=1} f_i \log^{}_2 f_i.$
The expected information gain is the change in entropy $H$ from a prior state $T$ to a state that takes some
information $a$ as given: 
\[ IG(T,a) = H(T) - H(T|a). \]
In this case, the information gain is IG = Entropy(parent node) - Weighted Sum of Entropy(Children)
\item variable importance is the sum of IG over all notes and all trees for that variable.
\end{itemize}

@@

\subsubsection{Synthetic data} 
\marginpar{these are not actually encoded as factors}
Each dataset consists of $n$ samples and $m$ variables where $n << m$, and values for each variable are ordinal factors with
three levels represented as numbers $\{0, 1, 2\}$ (which correspond to an additive effect encoding of genomics variation) randomly
generated from uniform distribution with equal probabilities.  The dichotomous response is generated from a
Bernoulli trail with $p$ (probability of success) calculated by applying an inverse logit transformation to the continuous
linear response ($y$) of five variables:
$$y = a_{10}* v_{10} + a_{100}*v_{100} + a_{1000}*v_{1000} + a_{10000}*v_{10000} + a_{10000}*v_{100000}$$
where $a_{10} = 0.8, a_{100} = 0.9, a_{1000} = 1.1, a_{10000} = 1.0, a_{10000}=1.15$.
